Managing variability in software systems

### 2021-05-27 Paper.tex
add corrections

\begin{abstract}
With the growing complexity of software systems, variability of said has become an increasing problem and managing it is thus a crucial part of the present-day software engineering process. It is particularly vital if either many product versions make use of an existing architecture over a long period of time or if a variety of software systems share a common code base. The focus of this paper is to provide an overview of how variation in software systems is caused, how it influences the development process and how it can be dealt with.
\end{abstract}

@@

We will try to provide a basic introduction of how variability is managed in software systems by collecting information from other studies which describe the key aspects of this vast field of research. In the first section we are providing some base knowledge acquired previous studies. After that we will lay out how one can identify the various sources of variability in software systems. Then we will bring up techniques and processes which have the purpose of modeling the variation in a software system. Last but not least we will discuss some of the big tasks which have to be addressed and conclude the topic.

@@

There are also several reasons why such variations have to exist in the first place \cite{b3}:
\begin{enumerate}
	\item \textit{Variation in function.} Not every product supports the same functions. Take as example a modern smartphone, some might support fingerprint authentication, some might support facial recognition authentication and some support neither.\\
	\item \textit{Variation in data.} Products might vary in the definitions of data structures. For example a software managing planned repairs of industrial machines might need store additional information for some companies like an SAP id in case these planned repairs are also concurrently managed in SAP, while some other companies might not need that information.\\
	\item \textit{Variation in technology.} The technology different products run on may vary. Some software systems for example need to run on Linux while others need to support Windows. Or for some a simple MySQL database might suffice while others need the SQL Server Enterprise edition because the amount of data might vary significantly.\\
	\item \textit{Variation in environment.} The environment software exists in might vary. Meaning also the way in which it interacts with its environment. \\
\end{enumerate}
These, at least, are the four most common origins of variation. Of course some more do exist, but these are beyond the scope of this document.\\
\section{Variability modeling}
As mentioned, there are two main approaches for managing variability. For each of which we will discuss the advantages and disadvantages and create a broad overview over the different modeling approaches.\\\\\\
\textbf{Feature modeling}\\
Features in the context of feature modeling are behaviours, characteristics or functionalities which are observable by the user of the software system. They can be used to define commonalities and differences between systems and their variants, as well as determine mandatory, optional or alternative aspects of them\cite{b8}. A broadly used approach of displaying features and their relations is a hierarchically structured diagram which is called a features diagram.\\

@@

If a feature is mandatory it has to be selected if its parent feature is selected, the type of the parent feature does not matter. A optional feature can be selected freely but there may be constraints from composition rules (e.g. a feature can only be choosen as part of a feature sets). Alternative features are can be pictured as a choice box, exactly \textit{one} item must be chosen from a feature set.\\

@@

Some of them are using contraints, similar to the approach used in feature modeling. Another approach for modeling dependencies is with rules or decisions, which are are checked when a decision is taken.\cite{b11} \\

@@

Further explanation of these binding times can be found in the subsequent paragraphs:\\
\begin{itemize}
	\item \textit{Product architecture derivation.} The architecture of the product already binds to a variation. This is the lowest binding time and will influence all subsequent steps.\\
	\item \textit{Compilation.} When the source code is compiled there exist several options to bind to a variation by adjusting constants, macros or compiler directives.\\
	\item \textit{Linking.} This option does not exist for any technology, but for many programming languages the linking process can be manually changed. For example by switching the DLL of a module.\\
	\item \textit{Startup} A simple option to bind variants to the system is by configuration files that influence the features a software system exhibits.\\
	\item \textit{Adaptation to runtime environment} This typically refers to plugins of a software, which can add variation to a system and can either be provided by the original developer or even third party vendors. \\
\end{itemize}
Any mechanism to bind to none, one or many different variants tend to fall into different categories presented by \cite{b2} and \cite{b3}, and also have parallels to the feature types presented in the previous section:\\
\begin{itemize}
	\item \textit{Variant Entity.} There exist several options but only one may be bound to at any given time.\\
	\item \textit{Optional Entity.} As the name suggests, an optional entity may or may not be included in the system.\\
	\item \textit{Multiple Coexisting Entities.} Similar to the variant entity but with the exception that multiple variants can be used.\\
\end{itemize}
These categories tend to be the same for the different binding times, but of course vary greatly in their implementation. While an optional entity during compilation may only be influenced by a compiler flag, it might be a whole plugin from a third party vendor when adapting a given runtime environment. An in depth analysis of these mechanisms can be found in \cite{b2}.

@@

Another aspect is that feature modeling is in its core a simple concept. Even though its impact on the industry is enormous considering it is the most popular concept for variability modeling in software system as of now.\cite{b9} Due to the big amount of research having been done in this field we understand the subject reasonably well by now.\\

As mentioned in \cite{b4} the currently existing methods also have their limitations. 
\begin{itemize}
\item \textit{Guaranteed correctness and consistency}. Authors have to urge for verification and consistency checking.\\
\item \textit{Performance overheads} are difficult to model.\\
\item \textit{Runtime upgradability} and the process of switching features is sometimes necessary.\\
\item \textit{Lack of support for security needs} can be a factor whichs limits the use of feature modeling.\\
\item \textit{Poor user-friendliness} can be introduced by poor handling of variability, which decreases the usability of a system.\\
\end{itemize}

@@

It should also be noted most of the studies in the area of FM are including some sort of tool to support the process. Only a few don't mention any form of software support.\\

\section{Conclussion}
In this document we have given a short introduction to the origin of variability, how to model it and what patterns exist as well as look over work related to these topics. \\

>>>>

\begin{abstract}
With the growing complexity of software systems, variability of said systems has become an increasing problem and managing it is thus a crucial part of the present-day software engineering process. It is particularly vital if either many product versions make use of an existing architecture over a long period of time or if a variety of software systems share a common code base. The focus of this paper is to provide an overview of how variation in software systems is caused, how it influences the development process and how it can be dealt with.
\end{abstract}

@@

We will try to provide a basic introduction of how variability is managed in software systems by collecting information from other studies which describe the key aspects of this vast field of research. In the first section we are providing some base knowledge acquired from previous studies. After that we will lay out how one can identify the various sources of variability in software systems. Then we will bring up techniques and processes which have the purpose of modeling the variation in a software system. Last but not least we will discuss some of the big tasks which have to be addressed and conclude the topic.

@@

There are also several reasons why such variations have to exist in the first place \cite{b3}:
\begin{enumerate}
	\item \textit{Variation in function.} Not every product supports the same functions. Take as example a modern smartphone, some might support fingerprint authentication, some might support facial recognition authentication and some support neither.\\
	\item \textit{Variation in data.} Products might vary in the definitions of data structures. For example a software managing planned repairs of industrial machines might need store additional information for some companies like an SAP id in case these planned repairs are also concurrently managed in SAP, while some other companies might not need that information.\\
	\item \textit{Variation in technology.} The technology different products run on may vary. Some software systems for example need to run on Linux while others need to support Windows. Or for some a simple MySQL database might suffice while others need the SQL Server Enterprise edition because the amount of data might vary significantly.\\
	\item \textit{Variation in environment.} The environment software exists in might vary. Meaning also the way in which it interacts with its environment. \\
\end{enumerate}

\section{Variability modeling}
As mentioned, there are two main approaches for managing variability. For each of which we will discuss the advantages and disadvantages and create a broad overview over the different modeling approaches.\\\\\\
\textbf{Feature modeling}\\
Features in the context of feature modeling are behaviours, characteristics or functionalities which are observable by the user of the software system. They can be used to define commonalities and differences between systems and their variants, as well as determine mandatory, optional or alternative aspects of them\cite{b8}. A broadly used approach of displaying features and their relations is a hierarchically structured diagram which is called a features diagram.\\

@@

If a feature is mandatory it has to be selected if its parent feature is selected, the type of the parent feature does not matter. An optional feature can be selected freely but there may be constraints from composition rules (e.g. a feature can only be choosen as part of a feature sets). Alternative features are can be pictured as a choice box, exactly \textit{one} item must be chosen from a feature set.\\

@@

Some of them use contraints, similar to the approach used in feature modeling. Another approach for modeling dependencies is with rules or decisions, which are are checked when a decision is taken.\cite{b11} \\

@@

Further explanation of these binding times can be found in the subsequent paragraphs:\\
\begin{itemize}
	\item \textit{Product architecture derivation.} The architecture of the product already binds to a variation. This is the lowest binding time and will influence all subsequent steps.\\
	\item \textit{Compilation.} When the source code is compiled there exist several options to bind to a variation by adjusting constants, macros or compiler directives.\\
	\item \textit{Linking.} This option does not exist for any technology, but for many programming languages the linking process can be manually changed. For example by switching the DLL of a module.\\
	\item \textit{Startup} A simple option to bind variants to the system is by configuration files that influence the features a software system exhibits.\\
	\item \textit{Adaptation to runtime environment} This typically refers to plugins of a software, which can add variation to a system and can either be provided by the original developer or even third party vendors. \\
\end{itemize}
Any mechanism to bind to none, one or many different variants tend to fall into different categories presented by \cite{b2} and \cite{b3}, and also have parallels to the feature types presented in the previous section:\\
\begin{itemize}
	\item \textit{Variant Entity.} There exist several options but only one may be bound to at any given time.\\
	\item \textit{Optional Entity.} An optional entity may or may not be included in the system.\\
	\item \textit{Multiple Coexisting Entities.} Similar to the variant entity but with the exception that multiple variants can be used.\\
\end{itemize}
These categories tend to be the same for the different binding times, but of course vary greatly in their implementation. While an optional entity during compilation may only be influenced by a compiler flag, it might be a whole plugin from a third party vendor when adapting a given runtime environment. An in depth analysis of these mechanisms can be found in \cite{b2}.

@@

Another aspect is that feature modeling is in its core a simple concept. Even though its impact on the industry is enormous considering it is the most popular concept for variability modeling in software system as of now.\cite{b9} \\

As mentioned in \cite{b4} the currently existing methods also have their limitations. 
\begin{itemize}
\item \textit{Guaranteed correctness and consistency}. Authors have to urge for verification and consistency checking.\\
\item \textit{Performance overheads} are difficult to model.\\
\item \textit{Runtime upgradability} and the process of switching features is sometimes necessary.\\
\item \textit{Lack of support for security needs} can be a factor which limits the use of feature modeling.\\
\item \textit{Poor user-friendliness} can be introduced by poor handling of variability, which decreases the usability of a system.\\
\end{itemize}

@@

It should also be noted most of the studies in the area of FM include some sort of tool to support the process. Only a few don't mention any form of software support.\\

\section{Conclusion}
In this document we have given a short introduction to the origin of variability, how to model it and what patterns exist as well as look over work related to these topics. \\
